---
title: "CAWS21 predictions"
output: 
  flexdashboard::flex_dashboard:
    navbar:
      - {title: "Main Codex page", href: "/project/codex", align: right}
      - {title: "Main site", href: "/", align: right}
    theme: bootstrap
    css: tables.css
    self_contained: no
    lib_dir: libs
    mathjax: local
author: "Mark Webster"
date: 2022-02-06
---

```{r, "setup", include=FALSE}
current <- "CAWS21"
dropped_players <- character()
library(ggplot2)
grey <- "grey13"
white <- "white"
theme_codex <- theme_dark() +
  theme(
    panel.border = element_rect(fill = NA, colour = grey),
    plot.background = element_rect(fill = grey, colour = grey),
    axis.ticks = element_line(colour = white),
    strip.background = element_rect(fill = grey, colour = NA),
    legend.background = element_rect(fill = grey, colour = NA),
    text = element_text(colour = white),
    axis.text = element_text(colour = white), 
  )
theme_set(theme_codex)
library(data.table)
library(codexdata)
library(codex)
library(knitr)
library(plotly)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, "read model results"}
vs_results <- readRDS("website_current_vs_split.rds")
```

```{r, "current matchups"}
current_players <- decks[tournament == current, player]
which_dropped <- match(dropped_players, current_players)
nondropped_players <- if (length(which_dropped) == 0) {
  current_players
}else{
  current_players[-which_dropped]
}
current_decks <- decks[tournament == current, deck]
nondropped_decks <- if (length(which_dropped) == 0) {
  current_decks
}else{
  current_decks[-which_dropped]
}
current_entries <- data.table(
  player = current_players, 
  deck = current_decks, 
  stringsAsFactors = TRUE
)
current_valid <- !is.na(current_decks)
nondropped_valid <- !is.na(nondropped_decks)
```

```{r, "current matches"}
current_matches <- fread("website_current_matches.csv", na.strings = "")[
  tournament == current
][order(round, round_match_number)]
finished_current_matches <- current_matches[!is.na(victor)]
```

```{r, "current mean matchups"}
current_matchup_array <- get_matchup_array(
  vs_results, 
  current_decks[current_valid], 
  current_players[current_valid]
)
current_matchups <- colMeans(current_matchup_array)
```

```{r, "nondropped mean matchups"}
nondropped_deck_matchup_array <- get_matchup_array(
  vs_results,
  nondropped_decks[nondropped_valid]
)
nondropped_matchup_array <- get_matchup_array(
  vs_results,
  nondropped_decks[nondropped_valid],
  nondropped_players[nondropped_valid]
)
nondropped_matchups <- colMeans(nondropped_matchup_array)
```

```{r, "current mean Nash"}
nondropped_deck_nash_samples <- get_nash_equilibria(nondropped_deck_matchup_array)
nondropped_deck_mean_nash <- as.data.table(
  apply(nondropped_deck_nash_samples, 1:2, mean), 
  keep.rownames = "Player"
)[
  ,
  c(.(Player = factor(Player, c("P1", "P2", "Both"))), .SD), 
  .SDcols = dimnames(nondropped_deck_nash_samples)[[2]]
]
nondropped_nash_samples <- get_nash_equilibria(nondropped_matchup_array)
nondropped_mean_nash <- as.data.table(
  apply(nondropped_nash_samples, 1:2, mean), 
  keep.rownames = "Player"
)[,
  c(.(Player = factor(Player, c("P1", "P2", "Both"))), .SD), 
  .SDcols = dimnames(nondropped_nash_samples)[[2]]
]
```

```{r, "current matchup progression"}
current_match_predictions <- current_matches[
  ,
  .(
    round, round_match_number,
    player1, player2,
    deck1, deck2,
    victor = factor(
      ifelse(
        is.na(victor) | victor == "", "ongoing",
        ifelse(victor == player1, "P1", "P2")
      ),
      levels = c("P1", "ongoing", "P2")
    ),
    victory,
    outcome = factor(
      ifelse(
        is.na(victory) | victory == "normal" | victory == "",
        ifelse(
          is.na(victor) | victor == "",
          "ongoing",
          ifelse(victor == player1, "P1", "P2")
        ),
        victory
      ),
      levels = c("P1", "ongoing", "P2", "forfeit", "timeout")
    ),
    # `P1 win probability` = diag(current_matchups[player1, player2]))
    `P1 win probability` = mapply(
      function(p1, p2, d1, d2) {
        nondropped_matchups[
          p1, # p1:d1 goes here when players have multiple decks
          p2
        ]
      },
      player1,
      player2,
      deck1,
      deck2
    )
  )
][
  , c(.SD, .(fairness = 1 - 2*abs(`P1 win probability` - 1/2)))
][
  ,
  c(
    .SD,
    .(text = paste0(
      "round: ", round,
      "\nP1: ", paste(player1, deck1),
      "\nP2: ", paste(player2, deck2),
      "\nP1 win probability: ", round(`P1 win probability`, digits = 3),
      "\nfairness: ", round(fairness, digits = 3),
      "\nvictor: ",
      ifelse(
        victor == "P1",
        player1,
        ifelse(victor == "P2", player2, "ongoing")
      )
    ))
  )
]
```

```{r, "current prediction evaluation"}
if (nrow(current_match_predictions) > 0) {
  current_matches_result_tallies <- finished_current_matches[
    victory == "normal",
    .(
      P1 = as.factor(player1),
      P2 = as.factor(player2),
      `P1 win` = victor == player1
    )
  ]
  current_matches_counts <- current_matches_result_tallies[
    ,
    .(`P1 wins` = paste(sum(`P1 win`), .N, sep = "/")),
    by = c("P1", "P2")
  ]
  observed_overall_freq <- current_match_predictions[
    !is.na(victory) & victory == "normal",
    mean(victor == "P1")
  ]
  freq_by_pred <- current_match_predictions[
    !is.na(victory) & victory == "normal",
    .(
      N = .N,
      bin_res = .N*(mean(victor == "P1") - observed_overall_freq)^2,
      bin_rel = sum((`P1 win probability` - mean(victor == "P1"))^2)
    ),
    by = .(pred_group = round(`P1 win probability`, 1))
  ]
  brier_decomp <- freq_by_pred[
    ,
    .(
      resolution = sum(bin_res)/sum(N),
      reliability = sum(bin_rel)/sum(N)
    )
  ]
  prior_decomp <- current_match_predictions[
    !is.na(victory) & victory == "normal",
    .(
      uncertainty = mean(`P1 win probability`)*(1 - mean(`P1 win probability`)),
      resolution =
        mean(`P1 win probability`)*(1 - mean(`P1 win probability`)) -
        mean(`P1 win probability`*(1 - `P1 win probability`))
    )
  ]
  current_score <- current_match_predictions[
    !is.na(victory) & victory == "normal",
    .(
      forecast = c("always 5-5", "prior expected score", "model"),
      `Brier score` = c(
        0.25,
        mean(`P1 win probability`*(1 - `P1 win probability`)),
        mean(((victor == "P1") - `P1 win probability`)^2)
      ),
      # decomp for prior expected is wrong, should be calculating uncertainty from predictions
      uncertainty = c(
        observed_overall_freq*(1 - observed_overall_freq),
        prior_decomp$uncertainty,
        observed_overall_freq*(1 - observed_overall_freq)
      ),
      resolution = c(
        0,
        prior_decomp$resolution,
        brier_decomp$resolution
      ),
      reliability = c(
        (1/2 - observed_overall_freq)^2,
        0,
        brier_decomp$reliability
      )
    )
  ]
  current_matchups_table <- as.data.table(
    current_matchups,
    keep.rownames = "P1"
  )[,
    c(list(P1 = factor(P1, unique(P1))), .SD),
    .SDcols = setdiff(colnames(current_matchups), "P1")
    ] |>
    melt(id.vars = "P1", variable.name = "P2", value.name = "matchup")
  current_lognormal_eval <- current_matchups_table[
    current_matches_counts,
    on = c("P1", "P2")
    ][order(P1, P2)]
}
```

# Tournament & evaluation {data-navmenu="Predictions"}

## Plots {data-width=400}

###

```{r, "section: current matchup progression"}
if (nrow(current_match_predictions) > 0)
  (
    ggplot(
      current_match_predictions,
      aes(x = round, y = `P1 win probability`, colour = outcome, text = text)
    ) +
      geom_point() +
      ylim(0, 1) +
      scale_x_discrete(limits = factor(seq.int(max(current_match_predictions$round)))) +
      scale_colour_viridis_d(option = "B", drop = FALSE)
  ) |> 
  ggplotly(tooltip = c("text")) |>
  config(displayModeBar = FALSE) |> 
  partial_bundle()
```

`r if(nrow(current_match_predictions) == 0) "waiting for match results..."`

###

```{r, "section: current fairness progression"}
if (nrow(current_match_predictions) > 0)
  (
    ggplot(
      current_match_predictions,
      aes(x = round, y = `fairness`, colour = outcome, text = text)
    ) +
      geom_point() +
      ylim(0, 1) +
      scale_x_discrete(limits = factor(seq.int(max(current_match_predictions$round)))) +
      scale_colour_viridis_d(option = "B", drop = FALSE)
  ) |> 
  ggplotly(tooltip = c("text")) |>
  config(displayModeBar = FALSE) |>
  partial_bundle()
```

`r if(nrow(current_match_predictions) == 0) "waiting for match results..."`

## Evaluation for normal finished matches {data-width=600}

### {data-height=220}

```{r, "section: current scores"}
if (nrow(current_match_predictions) > 0)
  kable(current_score, digits = 3)
```

`r if(nrow(current_match_predictions) == 0) "waiting for match results..."`

### {data-height=780}

```{r, "section: current evaluation table"}
if (
  !anyDuplicated(current_players[current_valid]) &&
  nrow(current_match_predictions) > 0
) {
  kable(current_lognormal_eval[order(-matchup)], digits = 3)
}
```

`r if(nrow(current_match_predictions) == 0) "waiting for match results..."`

# Nash equilibria for entrants {data-orientation=rows data-navmenu="Predictions"}

Nash equilibrium is not a good indicator of who's most likely to win the tournament, but it's a quick, simple heuristic to find likely contenders. Weights given are averaged over Nash equilibria calculated for each simulation sample (n = 4000).

## Mean Nash win rate {data-height=300}

### Deck-only mean win rate

```{r, "section: deck mean Nash win"}
kable(print_nash(nondropped_deck_mean_nash), digits = 3)
```

### Overall mean win rate

```{r, "section: overall mean Nash win"}
kable(print_nash(nondropped_mean_nash), digits = 3)
```

## Mean Nash pick rate {data-height=700}

### Deck-only mean pick distribution

```{r, "section: deck mean Nash spread"}
kable(reformat_used_nash(nondropped_deck_mean_nash), digits = 3)
```

### Overall mean pick distribution

```{r, "section: overall mean Nash spread"}
kable(reformat_used_nash(nondropped_mean_nash), digits = 3)
```

# Entrants {data-navmenu="Data"}

```{r "section: entrants"}
kable(current_entries)
```

# Matches {data-navmenu="Data"}

```{r, "section: matches"}
kable(
  current_matches[
    ,
    .(
      round,
      player1, player2,
      deck1, deck2,
      victor = ifelse(is.na(victor), "", victor),
      victory = ifelse(is.na(victory), "", victory)
    )
  ],
  digits = 3)
```
